---
title: "H3 nucMACC score characterization"
author: "Sara Wernig-Zorc"
date: "16/09/2021"
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	error = FALSE,
	warning = FALSE)
```

```{r load R data}
load(file = "MNase-H3.RData")
```

### Load libraries
```{r libraries, include=FALSE}
library(dplyr)
library(LSD)
library(corrr)
library(GenomicRanges)
library(rtracklayer)
library(ChIPseeker)
library(TxDb.Dmelanogaster.UCSC.dm3.ensGene)
library(GenomicFeatures)
library(ggplot2)
library(clusterProfiler)
library(edgeR)
library(venneuler)
library(ChIPpeakAnno)
library(enrichplot)
library(gtools)
library(corrplot)
```

# Mono- nucleosomes
## Mono-nucs: % of hypo-/hyper- accessible nucs
```{r import mono-nucs}
mononucs <- read.table("/path/to/RUN/09_nucMACC/nucMACC_result_table.tsv",header=T,sep="\t") 

lowMACC_nucs <- read.table("/path/to/RUN/09_nucMACC/hypoAcc_monoNucs.bed",header=F,sep="\t") 
colnames(lowMACC_nucs) <- cbind("Chr","Start","End","NucID","NucMACC","Strand")

highMACC_nucs <- read.table("/path/to/RUN/09_nucMACC/hyperAcc_monoNucs.bed",header=F,sep="\t")
colnames(highMACC_nucs) <- cbind("Chr","Start","End","NucID","NucMACC","Strand")

mono_nucs <- GRanges(mononucs)

nrow(lowMACC_nucs) / nrow(mononucs) *100
nrow(lowMACC_nucs)

nrow(highMACC_nucs) / nrow(mononucs) *100
nrow(highMACC_nucs)

c.low.access <- GRanges(lowMACC_nucs)
c.high.access <- GRanges(highMACC_nucs)
```

## Mono-nucs: Feature distribution
```{r mononucs: plot annotation}
db <- TxDb.Dmelanogaster.UCSC.dm3.ensGene
promoter <- getPromoters(TxDb=db, upstream=1500, downstream=350)

mono.nucs <- list(all_nucMACC=mono_nucs,
                  low_nucMACC=c.low.access,
                  high_nucMACC=c.high.access)
mono.nucs.anno <- lapply(mono.nucs, annotatePeak, TxDb=db, tssRegion=c(-1000, 300), verbose=FALSE)

pdf("Mono-nucs_low-high-nucMACC.pdf")
plotAnnoBar(mono.nucs.anno)
dev.off()

print("Genomic Annotation Summary: all monoNucs")
mono.nucs.anno$all_nucMACC 
print("Genomic Annotation Summary: lowMACC monoNucs")
mono.nucs.anno$low_nucMACC
print("Genomic Annotation Summary: highMACC monoNucs")
mono.nucs.anno$high_nucMACC
```


## Sub-nucs: % of hypo-/hyper- stable nucs
```{r import sub-nucs}
subNucs <- read.table("/path/to/RUN/10_sub-nucMACC/sub-nucMACC_result_table.tsv",header=T,sep="\t") 

lowMACC_subNucs <- read.table("/path/to/RUN/10_sub-nucMACC/unStable_subNucs.bed",header=F,sep="\t") 
colnames(lowMACC_subNucs) <- cbind("Chr","Start","End","NucID","NucMACC","Strand")

highMACC_subNucs  <- read.table("/path/to/RUN/10_sub-nucMACC/nonCanonical_subNucs.bed",header=F,sep="\t")
colnames(highMACC_subNucs) <- cbind("Chr","Start","End","NucID","NucMACC","Strand")

sub_nucs <- GRanges(subNucs)

print("all subNucs")
nrow(subNucs)

print("lowMACC subNucs")
nrow(lowMACC_subNucs) / nrow(subNucs)*100
nrow(lowMACC_subNucs) 

print("highMACC subNucs")
nrow(highMACC_subNucs) / nrow(subNucs)*100
nrow(highMACC_subNucs) 

f.low.stability <- GRanges(lowMACC_subNucs) #
f.high.stability <- GRanges(highMACC_subNucs) #

db <- TxDb.Dmelanogaster.UCSC.dm3.ensGene
sub.nucs <- list(subNucs=sub_nucs,
                 fragNucs=f.low.stability,
                 altNucs=f.high.stability)
               
sub.nucs.anno <- lapply(sub.nucs, annotatePeak, TxDb=db,  tssRegion=c(-1000, 300), verbose=FALSE)
pdf("Sub-nucs_low-high-nucMACC.pdf")
plotAnnoBar(sub.nucs.anno)
dev.off()

print("Genomic Annotation Summary: unique subNucs")
sub.nucs.anno$subNucs 
print("Genomic Annotation Summary: non-canonical subNucs")
sub.nucs.anno$altNucs
print("Genomic Annotation Summary: un-stable/fragile subNucs")
sub.nucs.anno$fragNucs

```


## Mono-nucs: GO pathway enrichment analysis

```{r mono-nucs: Enrichment analysis}
################## high nucMACC mono-nucs ################## 
data <- as.data.frame(mono.nucs.anno$high_nucMACC)
gene_conversion <- bitr(data$geneId, fromType="FLYBASE", toType="SYMBOL", OrgDb="org.Dm.eg.db")
gene_symbols <- gene_conversion$SYMBOL

enrichedTerms_BP <- enrichGO(gene = gene_symbols,
                OrgDb          = org.Dm.eg.db,
                keyType        = "SYMBOL",
                ont            = "BP", # BP, MF, CC or ALL
                pAdjustMethod  = "BH",
                pvalueCutoff   = 0.01,
                qvalueCutoff   = 0.01)

enrichedTerms_MF <- enrichGO(gene = gene_symbols,
                OrgDb          = org.Dm.eg.db,
                keyType        = "SYMBOL",
                ont            = "MF", # BP, MF, CC or ALL
                pAdjustMethod  = "BH",
                pvalueCutoff   = 0.01,
                qvalueCutoff   = 0.01)

pdf("high-mono-nucMACC_Molecular-function.pdf",width=15, height=5)
dotplot(enrichedTerms_MF, showCategory=10)
dev.off()

pdf("high-mono-nucMACC_Biological-process.pdf",width=15, height=5)
dotplot(enrichedTerms_BP, showCategory=10)
dev.off()

################## low nucMACC mono-nucs ################## 

data <- as.data.frame(mono.nucs.anno$low_nucMACC)
gene_conversion <- bitr(data$geneId, fromType="FLYBASE", toType="SYMBOL", OrgDb="org.Dm.eg.db")
gene_symbols <- gene_conversion$SYMBOL

enrichedTerms_BP <- enrichGO(gene = gene_symbols,
                OrgDb          = org.Dm.eg.db,
                keyType        = "SYMBOL",
                ont            = "BP", # BP, MF, CC or ALL
                pAdjustMethod  = "BH",
                pvalueCutoff   = 0.01,
                qvalueCutoff   = 0.01)

enrichedTerms_MF <- enrichGO(gene = gene_symbols,
                OrgDb          = org.Dm.eg.db,
                keyType        = "SYMBOL",
                ont            = "MF", # BP, MF, CC or ALL
                pAdjustMethod  = "BH",
                pvalueCutoff   = 0.01,
                qvalueCutoff   = 0.01)

pdf("low-mono-nucMACC_Molecular-function.pdf",width=15, height=5)
dotplot(enrichedTerms_MF, showCategory=10)
dev.off()

pdf("low-mono-nucMACC_Biological-process.pdf",width=15, height=5)
dotplot(enrichedTerms_BP, showCategory=10)
dev.off()
```

## Sub-nucs: GO enrichment analysis
### low sub-nucMACC - Fragile (hypo-stability) nucleosomes

```{r Fragile sub-nucs which do not overlap with mono nucs}
##### ANNOTATION AND PATHWAY ENRICHMENT
db <- TxDb.Dmelanogaster.UCSC.dm3.ensGene
promoter <- getPromoters(TxDb=db, upstream=1500, downstream=300)

data <- as.data.frame(sub.nucs.anno$fragNucs)
gene_conversion <- bitr(data$geneId, fromType="FLYBASE", toType="SYMBOL", OrgDb="org.Dm.eg.db")
gene_symbols <- gene_conversion$SYMBOL

enrichedTerms_BP <- enrichGO(gene = gene_symbols,
                OrgDb          = org.Dm.eg.db,
                keyType        = "SYMBOL",
                ont            = "BP", # BP, MF, CC or ALL
                pAdjustMethod  = "BH",
                pvalueCutoff   = 0.01,
                qvalueCutoff   = 0.01)

enrichedTerms_MF <- enrichGO(gene = gene_symbols,
                OrgDb          = org.Dm.eg.db,
                keyType        = "SYMBOL",
                ont            = "MF", # BP, MF, CC or ALL
                pAdjustMethod  = "BH",
                pvalueCutoff   = 0.01,
                qvalueCutoff   = 0.01)

pdf("S2_H3-FragileNucs_enriched-MolFuction.pdf")
dotplot(enrichedTerms_MF, showCategory=10)
dev.off()

pdf("S2_H3-FragileNucs_enriched-BioProcess.pdf")
dotplot(enrichedTerms_BP, showCategory=10)
dev.off()

write.table(data, file="fragNucs_annotated.csv",sep="\t",quote=FALSE)
```

## Unique subNucs - high nucMACC - non-canonical structres (Group 2)
```{r Group2: low access. sub nucs which do not overlap with mono nucs}
############# ANNOTATION AND PATHWAY ENRICHMENT
data <- as.data.frame(sub.nucs.anno$altNucs)
gene_conversion <- bitr(data$geneId, fromType="FLYBASE", toType="SYMBOL", OrgDb="org.Dm.eg.db")
gene_symbols <- gene_conversion$SYMBOL

enrichedTerms_BP <- enrichGO(gene = gene_symbols,
                OrgDb          = org.Dm.eg.db,
                keyType        = "SYMBOL",
                ont            = "BP", # BP, MF, CC or ALL
                pAdjustMethod  = "BH",
                pvalueCutoff   = 0.01,
                qvalueCutoff   = 0.01)

enrichedTerms_MF <- enrichGO(gene = gene_symbols,
                OrgDb          = org.Dm.eg.db,
                keyType        = "SYMBOL",
                ont            = "MF", # BP, MF, CC or ALL
                pAdjustMethod  = "BH",
                pvalueCutoff   = 0.01,
                qvalueCutoff   = 0.01)


pdf("S2_H3-AltNucs_enriched-MolFuction.pdf")
dotplot(enrichedTerms_MF, showCategory=10)
dev.off()

pdf("S2_H3-AltNucs_enriched-BioProcess.pdf")
dotplot(enrichedTerms_BP, showCategory=10)
dev.off()

write.table(data, file="altNucs_annotated.csv",sep="\t",quote=FALSE)
```

## nucMACC vs. Expression correlation
```{r nucs: corr with expression, echo=TRUE}
readCounts <- read.table("/Volumes/Sara_RAID_20TB/Sara_data_2018/MNase_project/MACC_project/results/RNA-seq/readCount/readCount_Ensembl_exon.txt",header=TRUE)

x <- DGEList(counts=readCounts[7:8])      #Makes a list for edgeR
normRead <- rpkm(x,gene.length = readCounts$Length)
expression <- rowMeans(normRead)
average <- cbind(readCounts,expression)

colnames(readCounts)[7] <- "rep1"
colnames(readCounts)[8] <- "rep2"

RPKM <- na.omit(average) 

#Cutoff based on RPKM density
pdf("RPKM_cut-off.pdf")
plot(density(log2(RPKM$expression)))
abline(v=1,col="blue",lwd=4)
dev.off()

expressed <- filter(RPKM, expression >= 1)
silent <- filter(RPKM, expression < 1)

####################################
low_monoNucs <- as.data.frame(mono.nucs.anno$low_nucMACC)
high_monoNucs <- as.data.frame(mono.nucs.anno$high_nucMACC)

unique_subNucs <- as.data.frame(sub.nucs.anno$unique_subNucs)
fragNucs <- as.data.frame(sub.nucs.anno$fragNucs)
altNucs <- as.data.frame(sub.nucs.anno$altNucs)

low_monoNucs.in.promoter <- filter(low_monoNucs,  annotation == "Promoter")
low_monoNucs.not.promoter <- filter(low_monoNucs,  annotation != "Promoter")
high_monoNucs.in.promoter <- filter(high_monoNucs,  annotation == "Promoter")
high_monoNucs.not.promoter <- filter(high_monoNucs,  annotation != "Promoter")

fragNucs.in.promoter <- filter(fragNucs,  annotation == "Promoter")
fragNucs.not.promoter <- filter(fragNucs,  annotation != "Promoter")
nrow(fragNucs.in.promoter)
nrow(fragNucs.not.promoter)

altNucs.in.promoter <- filter(altNucs,  annotation == "Promoter")
altNucs.not.promoter <- filter(altNucs,  annotation != "Promoter")
nrow(altNucs.in.promoter)
nrow(altNucs.not.promoter)
#
nucMACC.high.rpkm <- merge(high_monoNucs, RPKM, by.x = "geneId", by.y = "Geneid")
nucMACC.low.rpkm <- merge(low_monoNucs, RPKM, by.x = "geneId", by.y = "Geneid")

fragNucs.rpkm <- merge(fragNucs, RPKM, by.x = "geneId", by.y = "Geneid")
altNucs.rpkm <- merge(altNucs, RPKM, by.x = "geneId", by.y = "Geneid")

#
fragNucs.promoter.rpkm <- merge(fragNucs.in.promoter, RPKM, by.x = "geneId", by.y = "Geneid")
fragNucs.Not.promoter.rpkm <- merge(fragNucs.not.promoter, RPKM, by.x = "geneId", by.y = "Geneid")
altNucs.promoter.rpkm <- merge(altNucs.in.promoter, RPKM, by.x = "geneId", by.y = "Geneid")
altNucs.Not.promoter.rpkm <- merge(altNucs.not.promoter, RPKM, by.x = "geneId", by.y = "Geneid")

nucMACC.high.promoter.rpkm <- merge(high_monoNucs.in.promoter, RPKM, by.x = "geneId", by.y = "Geneid")
nucMACC.high.Not.promoter.rpkm <- merge(high_monoNucs.not.promoter, RPKM, by.x = "geneId", by.y = "Geneid")
nucMACC.low.promoter.rpkm <- merge(low_monoNucs.in.promoter , RPKM, by.x = "geneId", by.y = "Geneid")
nucMACC.low.Not.promoter.rpkm <- merge(low_monoNucs.not.promoter , RPKM, by.x = "geneId", by.y = "Geneid")

#subNucs
fragNucs.rpkm.promoter.expressed <- filter(fragNucs.promoter.rpkm,expression >= 1)
fragNucs.rpkm.promoter.silent <- filter(fragNucs.promoter.rpkm, expression < 1)
fragNucs.rpkm.not.promoter.expressed <- filter(fragNucs.Not.promoter.rpkm,expression >= 1)
fragNucs.rpkm.not.promoter.silent <- filter(fragNucs.Not.promoter.rpkm, expression < 1)
#monoNucs
nucMACC.high.rpkm.expressed <- filter(nucMACC.high.rpkm,expression >= 1)
nucMACC.high.rpkm.silent <- filter(nucMACC.high.rpkm,expression < 1)
nucMACC.low.rpkm.expressed <- filter(nucMACC.low.rpkm,expression >= 1)
nucMACC.low.rpkm.silent <- filter(nucMACC.low.rpkm, expression < 1)

nucMACC.high.rpkm.other.expressed <- filter(nucMACC.high.Not.promoter.rpkm,expression >= 1)
nucMACC.high.rpkm.other.silent <- filter(nucMACC.high.Not.promoter.rpkm,expression < 1)
nucMACC.high.rpkm.pro.expressed <- filter(nucMACC.high.promoter.rpkm,expression >= 1)
nucMACC.high.rpkm.pro.silent <- filter(nucMACC.high.promoter.rpkm,expression < 1)

nucMACC.low.rpkm.other.expressed <- filter(nucMACC.low.Not.promoter.rpkm,expression >= 1)
nucMACC.low.rpkm.other.silent <- filter(nucMACC.low.Not.promoter.rpkm, expression < 1)
nucMACC.low.rpkm.pro.expressed <- filter(nucMACC.low.promoter.rpkm,expression >= 1)
nucMACC.low.rpkm.pro.silent <- filter(nucMACC.low.promoter.rpkm, expression < 1)


sub_expressed <- cbind("Promoter_fragNucs"=nrow(fragNucs.rpkm.promoter.expressed)/nrow(fragNucs.in.promoter)*100,
                      "Intergenic_fragNucs"=nrow(fragNucs.rpkm.not.promoter.expressed)/nrow(fragNucs.not.promoter)*100)
                      
mono_expressed <- cbind("monoNucs_high"=nrow(nucMACC.high.rpkm.expressed)/nrow(high_monoNucs)*100,
                      "monoNucs_low"=nrow(nucMACC.low.rpkm.expressed)/nrow(low_monoNucs)*100)

sub_silent <- cbind("Promoter_fragNucs"=nrow(fragNucs.rpkm.promoter.silent)/nrow(fragNucs.in.promoter)*100,
                   "Intergenic_fragNucs"=nrow(fragNucs.rpkm.not.promoter.silent)/nrow(fragNucs.not.promoter)*100)

mono_silent <- cbind("monoNucs_high"=nrow(nucMACC.high.rpkm.silent)/nrow(high_monoNucs)*100,
                   "monoNucs_low"=nrow(nucMACC.low.rpkm.silent)/nrow(low_monoNucs)*100)

numbers_sub <- rbind("expressed" = sub_expressed, "silent" = sub_silent)
rownames(numbers_sub) <- c("expressed","silent")

numbers_mono <- rbind("expressed" = mono_expressed, "silent" = mono_silent)
rownames(numbers_mono) <- c("expressed","silent")

pdf("Corr_RNAexpression_monoNucs.pdf")
my_bar <- barplot(numbers_mono,border=FALSE, main="")
text(my_bar, numbers_mono[1,], paste("n = ", numbers_mono[1,], sep="") ,cex=1) 
text(my_bar, numbers_mono[2,], paste("n = ", numbers_mono[2,], sep="") ,cex=1)
dev.off()

pdf("Corr_RNAexpression_fragNucs.pdf")
my_bar <- barplot(numbers_sub,border=FALSE, main="")
text(my_bar, numbers_sub[1,], paste("n = ", numbers_sub[1,], sep="") ,cex=1) 
text(my_bar, numbers_sub[2,], paste("n = ", numbers_sub[2,], sep="") ,cex=1)
dev.off()
```

### Violin plot
```{r boxplot}
expression_data <- cbind(log2(nucMACC.high.rpkm$expression)+1,log2(nucMACC.low.rpkm$expression)+1,log2(fragNucs.rpkm$expression)+1,log2(altNucs.rpkm$expression)+1)

colnames(expression_data) <- c("hyperAccess","hypoAccess","unStabile","nonCanonical")

library(reshape2)
expr.reshape <- melt(expression_data)
boxplot <- ggplot(expr.reshape,aes(x=Var2,y=value,fill=Var2))

pdf("MonoNucs_hypo-hyper-access_corrExpr_boxplot.pdf")
boxplot + geom_violin(scale="area",trim=FALSE) +
  scale_y_continuous(limits = c(-10, 15)) + scale_fill_brewer(palette="YiGn") + geom_boxplot(width=0.1)
dev.off()

#FN sub-division
FN_data <- cbind(log2(fragNucs.promoter.rpkm$expression)+1,log2(fragNucs.Not.promoter.rpkm$expression)+1)
colnames(FN_data) <- c("promoter","other")

FN.reshape <- melt(FN_data)
FN.boxplot <- ggplot(FN.reshape,aes(x=Var2,y=value,fill=Var2))

pdf("SubNucs_hypo-stability_corrExpr_boxplot.pdf")
FN.boxplot + geom_violin(scale="area",trim=FALSE) +
  scale_y_continuous(limits = c(-10, 15)) + scale_fill_brewer(palette="RdPu") + geom_boxplot(width=0.1)
dev.off()

#altNuc sub-division
altNucs_data <- cbind(log2(altNucs.promoter.rpkm$expression)+1,log2(altNucs.Not.promoter.rpkm$expression)+1)
colnames(altNucs_data) <- c("promoter","other")

AN.reshape <- melt(altNucs_data)
AN.boxplot <- ggplot(AN.reshape,aes(x=Var2,y=value,fill=Var2))

pdf("SubNucs_hyper-stability_corrExpr_boxplot.pdf")
AN.boxplot + geom_violin(scale="area",trim=FALSE) +
  scale_y_continuous(limits = c(-10, 15)) + scale_fill_brewer(palette="RdPu") + geom_boxplot(width=0.1)
dev.off()

```


```{r save R data}
save.image(file = "MNase-H3.RData")
```

